Make any stupid plot you want

### 2020-12-31 draft/sections/abstract.tex
revised abstract

The Python Matplotlib visualization library grew organically such that the visualization functions are deeply entangled with the data the functions expect in such a way that it has become hard to adapt the library for modern needs. To rearchtecture the library while not sacrificing its current very high degree of flexiblity, we developed a model that preserves the structure of the data while also putting very few constraints on the data it can encode.

We propose that afunction that converts data into a graphic is the composition of two functions: a set of equivariant maps from data to visual channels and an equivariant map that assembles the visual channels into a graphic renderable by a physical display. We encode data as a fiber bundle to separately consider the fields of the data (the variables) as the fiber and the connectivity as the base space. We also represent prerender display spaces as a fiber bundle with a base space that is a continous map into the data base space, thereby preserving the connectivity of the data in the graphic.

We then illustrate this model through case studies of a scatter plot, line plot, and heatmap. Finally we show that this model can be realized in software with a small prototype. We encode the topological base spaces using triangulation, make use of programming types for the fiber, and build Matplotlib's existing infrastructure for the rendering.

In future work, we propose to develop the math further such that we can represent graphics that are composites of primitives and implement those ideas as user facing composite visualization functions.

>>>>

The difference between drawing and data visualization is that visualization implies that some structure of the data is preserved in the visual presentation. We formalize what structure is preserved and how via a mathematical topological artist model. In our model, data and graphics can be viewed as sections of fiber bundles; this allows for (1) decomposing the translation of data fields (variables) into visual channels via an equivariant map on the fibers and (2) a topology-preserving map of the base spaces that encode the connectivity of the points in the dataset and elements of the graphic. Furthermore, this model supports an algebraic sum operation such that more complex visualizations can be built from simple ones. We illustrate the application of the model through case studies of a scatter plot, line plot, and heatmap. We show that this model can be implemented with a small prototype.

We call the function mapping sections of the data bundle to the graphical bundle the "Artist" based on analogous architecture elements in the Python visualization library Matplotlib. To demonstrate the practical value of the model, we propose a model driven re-architecture of the artist layer of the Matplotlib. We represent the topological base spaces using triangulation, make use of programming types for the fiber, and build on Matplotlib's existing infrastructure for the rendering. In addition to providing a way to ensure the library preserves structure, the functional decomposition of the artist in the model could improve modularity, maintainability, and point to ways in which the library could better support concurrency and interactivity. The thesis will follow through on this proposal to explore how to further develop the model, showing how it can support Matplotlib's current diverse range of data visualizations while providing a better platform for domain-specific visualization library developers.
